Simulated Observations of Young Gravitationally Unstable Protoplanetary Discs

### 2013-05-07 young_pp_discs.tex
ref comments addressed

To appropriately describe the environment within which a young protoplanetary disc is embedded, we combine the disc model from I2011 with a model of a dense core with characteristics similar to the well-studied pre-stellar core L1544 (\citealt{Keto2010}, hereafter KC2010). Although L1544 does not appear to contain a central protostar and disc itself, it is very similar in structure to L1521F,  another dense core in Taurus which hosts a low mass protostar and possibly a disc \citep{Bourke2006}. The KC2010 model follows the dynamical, chemical and thermal evolution of a contracting Bonnor-Ebert sphere \citep{Bonnor1956,Ebert1957} with total mass of 10\,M$_{\odot}$, until it reaches the density, temperature and velocity profiles which best match observations. The pre-stellar core model adopted here contains slight modifications due to the inclusion of oxygen cooling in the outer regions of the cloud, where CO is mostly photodissociated \citep{Caselli2012}. Fig. \ref{fig:l1544_model} shows the physical parameters of the core model{\bf as used in an improved version of the KC2010 model (Keto et al. 2013, in preparation).}\smallskip

@@

The pre-stellar core structure is maintained down to a radius of 80\,au, within which the model of the young protoplanetary disc is used. The disc structure used is from the same hydrodynamics simulation described in I2011, which is a 1$\,$M$_\odot$ protostar surrounded by a 0.39$\,$M$_\odot$ disc. Fig.\ref{velocity} shows the azimuthally averaged rotational velocity in the disc plane, together with the best fit curve, which will be used in Sec.~\ref{sec:alma_predictions} to compare with position velocity diagrams. The system is envisaged to be at a very early stage of evolution and embedded in an envelope.  The still growing protostar will eventually become an A star through accretion of most of the disc's mass. The hydrodynamics simulation is run with a proper equation of state for molecular hydrogen \citep{Boley2007}.  The star is allowed to move freely in response to disc torques, and radiative cooling is included \citep{Boley2009}. The self-gravitating disc exhibits prominent spiral structure, with H$_2$ number densities in the disc ranging from 10$^4$-10$^{13}$ cm$^{-3}$ and temperatures from 30-400 K (fig. \ref{rhoT}). For the non-LTE 3D radiative transfer modelling presented here, we interpolate the I2011 simulation output on to a 256$\times$256$\times$64 Cartesian grid with spatial resolution of 0.5$\,$au in x, y and z and sample from this to create the unstructured radiative transfer grid (see section \ref{subsec:gridding}).  As in the hydrodynamics simulation, the dust and gas temperatures are assumed to be in equilibrium within the disc and a gas to dust mass ratio of 100 is used throughout the entirety of the model. The dust opacities were adopted from \citet{Ossenkopf1994} and refer to dust grains with thick icy mantles and 10$^6$ yr coagulation history.  Although dust grains within the protoplanetary disc are expected to coagulate further,  significantly affecting the opacity, we did not take grain growth into account in our model.  The dust opacities used in the radiative transfer are not the same as the ones used in the hydrodynamic model (which used the dust opacities of \citealt{DAlessio2001}). However, the results described here do not depend on the details of the dust opacity used during the hydrodynamic evolution.\smallskip

\subsection{Chemical structure} \label{subsec:chemical_structure}
Chemical abundances in the disc were taken from I2011, who followed gas-grain chemical processes during the dynamical evolution of the disc. As described in I2011, the hydrodynamical simulation included Lagrangian tracer fluid elements in addition to the solution of the equations of hydrodynamics on an Eulerian grid.  These fluid elements were used to record the thermal history of the gas as material passed through and between the spiral structure. I2011 used this time evolution information from fluid elements to calculate the abundances of 125 species, related by 1334 reactions.  These abundances were then interpolated by I2011 on to a Cartesian grid with cell sizes 2.2$\times$2.2$\times$0.22$\,$au$^3$ in x, y and z to reveal the morphology of the disc as seen by different molecular species. From these we selected the four species which appear to trace different regions of the disc: the inner 30\,au (OCS), the inner 40\,au (H$_2$CO),  the region between $\simeq$40 and 60\,au (HCO$^+$) and the entirety of the disc (C$^{17}$O) (see fig. \ref{Chemistry}). The abundance model used for C$^{17}$O was the CO model reduced by a factor of 1792, the ratio of $^{16}$O to $^{17}$O in the local ISM \citep{Wilson1994}. As explained by I2011,  H$_2$CO and OCS mostly probe the central warm regions, where icy mantles evaporate, whereas HCO$^+$ preferentially traces the outer spiral pattern as in the central region it is destroyed by water molecules and transformed into H$_3$O$^+$ and CO. The simple chemistry in the KC2010 model, adopted here as the envelope of the protoplanetary disc, does not provide detailed abundances of molecular species (besides CO and H$_2$O, see also \citealt{Caselli2012}). As discussed in the results section \ref{sec:model_results}, abundances have been adopted based on values measured toward similar objects. The molecular data used here for the radiative transfer are taken from the Leiden Atomic and Molecular DAtabase (LAMDA) (\citealt{Schoier2005} http://home.strw.leidenuniv.nl/$\sim$moldata/). \smallskip

{\bf Whilst photochemistry was not included in I2011, it is unclear to what extent the radiation from the protostar will impact upon the disc, given that material infalling from the envelope and newly forming outflows will be obscuring the star. Column densities between upper layers of the disc and the protostar in the model used here are on the order of 10$^{23}$ cm$^{-2}$ (is this enough to block the protostars visible and UV?). This is in contrast to later stages of disc evolution where the envelope has been almost entirely cleared (e.g. Walsh et. al), in this case it is clear that the chemistry in the upper layers of the disc will be dominated by photo-chemical processes.}

\subsection{The radiative transfer code} \label{subsec:radiative_transfer_code}
The radiative transfer program used is LIME (LIne Modelling Engine; \citealt{Brinch2010}), which  calculates line intensities based on a weighted sample of randomly chosen points in a continuous 3D model. The method of selecting these points is given in section \ref{subsec:gridding}. At each of these points, the density of the main collision partner (H$_2$), gas and dust temperatures, velocity, molecular abundances and unresolved turbulent velocity are specified. These points are then smoothed by Lloyd's algorithm \citep{Lloyd1982} in order to minimise the variation in distance between points whilst keeping the same underlying distribution. These points are then connected by Delaunay triangulation and it is between the points connected by this method that photon are allowed to propagate (fig. \ref{grid}). The level populations of the selected molecules are calculated at each of these points from collisional and radiative (de)excitation and the local radiation field is calculated. This is repeated 20 times with the populations of each level converging towards a single value. This number of iterations is sufficient for the signal to noise ratio of the level populations (as defined in \citealt{Brinch2010}) to exceed 1000 for 99\% of the points, ensuring that the simulation has converged on a stable level population. After 20 iterations the model is ray-traced in order to produce synthetic brightness maps. The average of ten separate runs was taken to minimise the artefacts in the output images, resulting from the grid construction (Fig. \ref{averages}).

@@

\subsection{Grid construction} \label{subsec:gridding}
In order to construct the grid, candidate points are randomly selected from the volume to be simulated. These candidates then have their H$_2$ number density, and the number density of the molecular species being simulated, compared against those of a reference point in order to decide if the candidate point is to be used in the grid or not. Candidate grid points are selected at random in a cylindrical coordinates that is linearly spaced in z and $\theta$ and logarithmically spaced in r. For each point to be selected, a random number $\alpha$ is drawn from the semi-open set [0,$\,$1) as a threshold. After selection of random coordinates, the hydrogen density and molecular density at the candidate point (n and m, respectively) are compared against the densities of a reference point on the inner edge of the disc (n$_0$ and m$_0$). If $\alpha<\left( \frac{n}{n_0} \right)^{0.3}$ or $\alpha< \left( \frac{m}{m_0} \right)^{0.3}$ then the point is selected for use. Otherwise another r, $\theta$, z co-ordinate is selected and it becomes the candidate point. 20\% of these selected points are forced to be at radii greater than $\sqrt{R_{min}R_{max}}$ (where $R_{min}$ and $R_{max}$ are the inner and outer radii of the model) in order to stop too many of the selected points clustering in the high density disc and leaving the envelope under sampled. In addition to this method of selection, 5\% of the points are linearly distributed in x, y and z with no bias with regards to density or abundance. This provides a minimum level of sampling for the large low density regions in the outer parts of the simulated volume. See fig. \ref{points} for an example of the points distribution in r, z. The function comparing the candidate point to the reference point and the candidate point distribution were selected empirically to sample all scales while ensuring that the majority of points are located in the inner disc where the density is higher.  \smallskip

@@

We focus on the frequency range available with ALMA, with particular attention to band 7, which offers the best trade off between resolution and sensitivity. Thus, we consider C$^{17}$O(3$\rightarrow$2), HCO$^+$(3$\rightarrow$2), OCS (26$\rightarrow$25) and H$_2$CO(4$_{04}\rightarrow$3$_{03}$).  The frequencies and upper level energies of these transitions are given in table \ref{sigmas}. The results presented in this section are limited to those lines which show detectable emission/absorption which can be used to trace either spiral structure or rotation.\smallskip

To simulate observations, the model was placed at roughly the distance of nearby low-mass star forming regions (100$\,$pc). The inclination angle was varied from 15$^\circ$ to 75$^\circ$ relative to the edge on case. From these simulated observations, integrated intensity maps, intensity weighted velocity maps and position velocity diagrams were created. The integrated intensity and  intensity weighted velocity maps (figs. \ref{sim_all}, and \ref{mom0_maps}) were from -12.5 to -0.5$\,$km$\,$s$^{-1}$ and +0.5 to +12.5$\,$km$\,$s$^{-1}$ to avoid domination by the contribution from the envelope, which can be seen in some PV diagrams as the strong absorption feature at all positions around zero velocity (see section \ref{sec:alma_predictions}). Intensity weighted velocity maps are shown with a cut-off of 3$\sigma$ as described in section \ref{sec:alma_predictions}.\smallskip

@@

OCS 26$\rightarrow$25 traces only the innermost 20$\,$au of the disc and can be used to examine the central regions (fig. \ref{sim_all}). As OCS is not seen in outflows (e. g. \citealt{Stanke2007}, \citealt{VDTak2003}), it can be used to measure the rotation of the central part of the disc without contamination from shocked material along the outflow. The OCS lines are unique amongst the lines simulated in that they trace radii $<$16$\,$au, even smaller than the extent of the OCS column density distribution shown in fig. \ref{Chemistry}. This is because the OCS line, which has the highest upper energy level among the selected transitions, traces only the hottest and densest regions of the disc where the high energy states are populated. As a result, the innermost spiral structure is best seen in OCS (26$\rightarrow$25), as shown in fig. \ref{sim_all}. The majority of the OCS line is visible in absorption against the bright continuum emission from the disc mid-plane. The exception to this is towards the centre of the disc where the hole in the hot, dense mid-plane means that there is little continuum to be absorbed. However, this may be an artefact of our treatment of the inner hole, given that we neglect the presence of the protostar and other hot material within the central 2$\,$au. Thus, we expect absorption also to be observed toward the central region of the disc.\smallskip

The abundance of the H$_2$CO molecule traces the spiral structure of the inner $\sim$40$\,$au of the disc, which can be seen in the integrated intensity map, and its rotation is detected out to larger radii compared to the OCS line (fig. \ref{sim_all}). As with the OCS line, the majority of the H$_2$CO 4$_{04}$$\rightarrow$3$_{03}$, is seen in absorption against the disc mid-plane continuum, with the central region seen in emission. However, unlike the OCS, the H$_2$CO extends out to large enough radii to trace the voids between the outer spiral arms. In these regions, the H$_2$CO line can be seen in emission.\smallskip

@@

Simulations at inclinations other than 30$^\circ$ were also performed with C$^{17}$O. Fig. \ref{other_inc} shows the C$^{17}$O 3$\rightarrow$2 line in the same physical model but at 15$^\circ$, 30$^\circ$, 45$^\circ$ and 75$^\circ$ to edge on. At 15$^\circ$ to edge on we can still see some structure in the integrated intensity map and the structure is also visible in the position velocity diagram. At larger inclinations it gets more difficult to obtain any kinematic information but the physical structure of the disc becomes clearer.

@@

\begin{figure*}
 \includegraphics[width=168mm]{Figures/sim/fig11.eps} 
 \caption{H$_2$CO 4$_{04}\rightarrow$3$_{03}$ channel maps as simulated with CASA. The contours start at 0.35 mJy$\,$beam$^{-1}\,$km$\,$s$^{-1}$  and are in steps of 0.35 mJy$\,$beam$^{-1}\,$km$\,$s$^{-1}$ (absorption in blue, emission in red). Overlaid on the 1mm continuum emission. the integration range for each map is 2.4 km$\,$s$^{-1}$, for the top left the range is from -10.8 to -8.4 km$\,$s$^{-1}$ and for the bottom right the range is from 8.4 to 10.8 km$\,$s$^{-1}$}
 \label{h2co_chanmap}
\end{figure*}

Fig. \ref{continuum} shows the 1\,mm continuum emission and beam size in the chosen ALMA configuration. The spiral structure of the disc is clearly visible at this inclination and and it becomes clearer at angles closer to face-on. A mass estimate for the disc was calculated using the equation $M=g S_\nu d^2/\kappa_\nu B_\nu(T_d)$ where g is the gas/dust mass ratio S$_\nu$ is the integrated flux, d is the distance to the source, $\kappa_\nu$ is the dust opacity and B$_\nu(T_d)$ is the black body function for a given dust temperature \citep{Beltran2006}. From this, a disc mass measurement of 0.1$\,$M$_\odot$ was obtained by assuming a dust temperature of 40K, and a measurement of 0.005$\,$M$_\odot$ was obtained by using 400K. The flux was integrated from y=0.2 to -0.2 and x=0.4 to -0.4. This shows that the disc is not optically thin to millimetre wavelengths. This underestimate of the disc mass by at least a factor of 4 shows that in young, massive discs such as those we are investigating, longer wavelength continuum observations are necessary for determining the disc mass.\smallskip

Fig. \ref{h2co_chanmap} shows a channel map of the H$_2$CO 4$_{04}\rightarrow$3$_{03}$ line overlaid on the 1$\,$mm continuum emission. From this it can be seen that the absorption features trace the disc spiral structure whilst the line emission emanates from gaps in the outer disc, where there is less continuum to be absorbed. The central panel is dominated by the envelope contribution, causing strong absorption through out the disc. The other panels show the line following a rotation profile distorted by the presence of the spiral arms. \smallskip

Fig. \ref{15deg} shows the C$^{17}$O 3$\rightarrow$2 integrated intensity and PV diagram for an inclination angle of 15$^\circ$ to edge on. Spiral structure can be clearly seen in the position velocity diagram even at low inclinations. The ``finger-like'' structures in the PV diagram are similar to those seen in observations of the Milky Way and in simulations of spirals in the Milky Way (e. g. \citealt{Bissantz2003}). Although the detail of the spiral structure (number of arms, pitch angle etc.) is difficult to gauge from such a PV diagram, high spectral and angular resolution observations can definitely unveil their presence and thus be used to find gravitationally unstable discs. \smallskip

@@

The spectra presented in Fig. \ref{spectra} were extracted from the maps in fig. \ref{mom0_maps} for the positions indicated by the corresponding coloured dots along the Y=0 axis. As with all the CASA simulated images, these spectra were simulated with an ALMA window with total bandwidth 938$\,$MHz, giving velocity resolutions of between 418 and 527\,m\,s$^{-1}$. Very notably, these spectral features are not symmetric about the disc centre. For example, C$^{17}$O(3--2) has a significantly stronger blueshifted absorption, due to the presence of a spiral arm along the same line of sight. In the OCS lines, the two sides are not moving with equal velocities and the line widths and shapes are markedly different. These can be attributed to the non-axisymmetric nature of the model.

@@

In this paper we have presented radiative transfer simulations of a hybrid model comprising a 0.39$\, M_\odot$ self gravitating disc with a radius of 64$\,$au and spiral density waves, surrounded by an envelope that is a contracting 10$\,M_\odot$ BE-sphere. The main results of these simulations are as follows:\newline
$\bullet$ CASA simulations of this model show that at a distance of 100$\,$pc, extraction of kinematic and structural information from the continuum and molecular lines is possible with ALMA band 7.\\
$\bullet$ Our simulations show that molecular features are predominantly seen in absorption against the hot mid-plane of self-gravitating protoplanetary discs, with emission only being seen where voids in the structure of the disc do not provide a bright continuum source to be absorbed.\\
$\bullet$ The quiescent nature of the envelope around such discs only affect lines within $\pm\,$0.5 km$\,$s$^{-1}$ of the LSR velocity.\\
$\bullet$ The lines studied (OCS 26$\rightarrow$25, H$_2$CO 4$_{04}$$\rightarrow$3$_{03}$, C$^{17}$O 3$\rightarrow$2 and HCO$^+$ 3$\rightarrow$2) taken together allow all regions of the disc to be sampled and the rotation of the disc to be constrained from the inner edge to the outer edge.\\
$\bullet$ Spiral structures in young embedded discs are detectable at a wide range of inclination angles, they can be resolved spatially at angles close to face on and can be inferred from position velocity diagrams at low inclination angles.\smallskip

@@

{\bf The disc model used does not include magnetic fields, including magnetic fields may suppress GI and spiral wave formation by slowing down or even preventing early disc growth in the case of sufficient magnetisation and alignment between the parent cores rotation axis and magnetic field direction (Joos et al. 2012, Li et al. 2013). This would mean that if observations similar to the ones we have described are made constraints could be placed on B-field strength within cores/discs. \smallskip Futher work to improve these simulated observations would include looking at the effects of material from the envelope falling onto the disc and how this affects the chemistry and line propagation, and including realistic outflows into the model.}

These results demonstrate that ALMA will be able to probe the very earliest stages of circumstellar discs which feed the growing protostar and eventually will evolve in a planetary system. Our results show that ALMA observations in band 7 will allow the assessment of whether or not these young discs are gravitationally unstable and test theories of pre-stellar to protostellar core evolution. If such gravitational instabilities exist they will have significant implications for the evolution of the disc. For example the will affect fragmentation, planetesimal formation \citep{Boley2009,Johnson2013,Gibbons2012} and the chemical evolution in the disc (I2011).

>>>>

To appropriately describe the environment within which a young protoplanetary disc is embedded, we combine the disc model from I2011 with a model of a dense core with characteristics similar to the well-studied pre-stellar core L1544 (\citealt{Keto2010}, hereafter KC2010). Although L1544 does not appear to contain a central protostar and disc itself, it is very similar in structure to L1521F,  another dense core in Taurus which hosts a low mass protostar and possibly a disc \citep{Bourke2006}. The KC2010 model follows the dynamical, chemical and thermal evolution of a contracting Bonnor-Ebert sphere \citep{Bonnor1956,Ebert1957} with total mass of 10\,M$_{\odot}$, until it reaches the density, temperature and velocity profiles which best match observations. The pre-stellar core model adopted here contains slight modifications compared to KC2010 due to the inclusion of oxygen cooling in the outer regions of the cloud (Keto et al. 2013, in preparation), where CO is mostly photodissociated \citep{Caselli2012}. Fig. \ref{fig:l1544_model} shows the physical parameters of the core model adopted here. \smallskip

@@

The pre-stellar core structure is maintained down to a radius of 80\,au, within which the model of the young protoplanetary disc is used. The disc structure is taken from the same hydrodynamics simulation of a 1$\,$M$_\odot$ protostar surrounded by a 0.39$\,$M$_\odot$ disc, described in I2011. Fig.\ref{velocity} shows the azimuthally averaged rotational velocity in the disc plane, together with the best fit curve, which will be used in Sec.~\ref{sec:alma_predictions} to compare with position velocity diagrams. The system is envisaged to be at a very early stage of evolution and embedded in an envelope.  The still growing protostar will eventually become an A star through accretion of most of the disc's mass. The hydrodynamics simulation is run with a proper equation of state for molecular hydrogen \citep{Boley2007}.  The protostar is allowed to move freely in response to disc torques, and radiative cooling is included \citep{Boley2009}. The self-gravitating disc exhibits prominent spiral structure, with H$_2$ number densities in the disc ranging from 10$^4$-10$^{13}$ cm$^{-3}$ and temperatures from 30-400 K (fig. \ref{rhoT}). {\bf The disc model does not include magnetic fields, which may suppress gravitational instability and spiral wave formation by slowing down or even preventing early disc growth in the case of sufficient magnetisation and alignment between the parent cores rotation axis and magnetic field direction \citep{Joos2012,Li2013}. Thus, the observation of spiral structure would provide strong constraints on the dynamical state of discs. \smallskip }

For the non-LTE 3D radiative transfer modelling presented here, we interpolate the I2011 simulation output on to a 256$\times$256$\times$64 Cartesian grid with spatial resolution of 0.5$\,$au in x, y and z and sample from this to create the unstructured radiative transfer grid (see section \ref{subsec:gridding}).  As in the hydrodynamics simulation, the dust and gas temperatures are assumed to be in equilibrium within the disc and a gas to dust mass ratio of 100 is used throughout the entirety of the model. The dust opacities were adopted from \citet{Ossenkopf1994} and refer to dust grains with thick icy mantles and a 10$^6$ yr coagulation history.  Although dust grains within the protoplanetary disc are expected to coagulate further,  significantly affecting the opacity, we did not take grain growth into account in our model.  The dust opacities used in the radiative transfer are not the same as the ones used in the hydrodynamic model (which used the dust opacities of \citealt{DAlessio2001}). However, the results described here do not depend on the details of the dust opacity used during the hydrodynamic evolution.\smallskip

\subsection{Chemical structure} \label{subsec:chemical_structure}
Chemical abundances in the disc were taken from I2011, who followed gas-grain chemical processes during the dynamical evolution of the disc. {\bf Photochemistry is not taken into account, based on the assumption that in the early  (Class 0, early ClassI) stages of protostellar accretion and evolution considered here, the protostar/disc system is heavily embedded in a thick envelope of gas and dust. The combination of envelope infall onto the disc and powerful outflows (see e.g. \citealt{Machida2013}) may drastically limit the illumination of the upper layers of young discs by the central protostar. Observations are needed to shed light on the importance of photochemistry at these early stages.}

As described in I2011, the hydrodynamical simulation included Lagrangian tracer fluid elements in addition to the solution of the equations of hydrodynamics on an Eulerian grid.  These fluid elements were used to record the thermal history of the gas as material passed through and between the spiral structure. I2011 used this time evolution information from fluid elements to calculate the abundances of 125 species, related by 1334 reactions.  These abundances were then interpolated by I2011 on to a Cartesian grid with cell sizes 2.2$\times$2.2$\times$0.22$\,$au$^3$ in x, y and z to reveal the morphology of the disc as seen by different molecular species. From these we selected the four species which appear to trace different regions of the disc: the inner 30\,au (OCS), the inner 40\,au (H$_2$CO),  the region between $\simeq$40 and 60\,au (HCO$^+$) and the entirety of the disc (C$^{17}$O) (see fig. \ref{Chemistry}). The abundance model used for C$^{17}$O was the CO model reduced by a factor of 1792, the ratio of $^{16}$O to $^{17}$O in the local ISM \citep{Wilson1994}. As explained by I2011,  H$_2$CO and OCS mostly probe the central warm regions, where icy mantles evaporate, whereas HCO$^+$ preferentially traces the outer spiral pattern as in the central region it is destroyed by water molecules and transformed into H$_3$O$^+$ and CO. The simple chemistry in the KC2010 model, adopted here as the envelope of the protoplanetary disc, does not provide detailed abundances of molecular species (besides CO and H$_2$O, see also \citealt{Caselli2012}). As discussed in the results section \ref{sec:model_results}, abundances in the envelope have been adopted based on values measured toward similar objects. The molecular data used here for the radiative transfer are taken from the Leiden Atomic and Molecular DAtabase (LAMDA) (\citealt{Schoier2005} http://home.strw.leidenuniv.nl/$\sim$moldata/). \smallskip

\subsection{The radiative transfer code} \label{subsec:radiative_transfer_code}
The radiative transfer program used is LIME (LIne Modelling Engine; \citealt{Brinch2010}), which  calculates line intensities based on a weighted sample of randomly chosen points in a continuous 3D model. The method of selecting these points is given in section \ref{subsec:gridding}. At each of these points, the density of the main collision partner (H$_2$), gas and dust temperatures, velocity, molecular abundances and unresolved turbulent velocity are specified. These points are then smoothed by Lloyd's algorithm \citep{Lloyd1982} in order to minimise the variation in distance between points whilst keeping the same underlying distribution. These points are then connected by Delaunay triangulation and it is between the points connected by this method that photons are allowed to propagate (fig. \ref{grid}). The level populations of the selected molecules are calculated at each of these points from collisional and radiative (de)excitation and the local radiation field is calculated. This is repeated 20 times with the populations of each level converging towards a single value. This number of iterations is sufficient for the signal to noise ratio of the level populations (as defined in \citealt{Brinch2010}) to exceed 1000 for 99\% of the points, ensuring that the simulation has converged on a stable level population. After 20 iterations the model is ray-traced in order to produce synthetic brightness maps. The average of ten separate runs was taken to minimise the artefacts in the output images, resulting from the grid construction (Fig. \ref{averages}).

@@

\subsection{Grid construction} \label{subsec:gridding}
{\bf In order to construct the grid, candidate points are randomly selected from the volume to be simulated. These candidate points are associated with values of the H$_2$ number density and the number density of the molecular species being simulated at the position of the point in the underlying model. These are then compared against the H$_2$ and molecular density of a reference point in order to decide if they will be included in the grid.

\bf Candidate grid points are selected at random in a cylindrical coordinate system that is linearly spaced in z and $\theta$ and logarithmically spaced in r. For each point to be selected, a random number $\alpha$ is drawn from the semi-open set [0,$\,$1) to be used as a threshold.} After selection of random coordinates, the hydrogen density and molecular density at the candidate point (n and m, respectively) are compared against the densities of a reference point on the inner edge of the disc (n$_0$ and m$_0$). If $\alpha<\left( \frac{n}{n_0} \right)^{0.3}$ or $\alpha< \left( \frac{m}{m_0} \right)^{0.3}$ then the point is selected for use. Otherwise another r, $\theta$, z co-ordinate is selected and it becomes the candidate point. 20\% of these selected points are forced to be at radii greater than $\sqrt{R_{min}R_{max}}$ (where $R_{min}$ and $R_{max}$ are the inner and outer radii of the model) in order to stop too many of the selected points clustering in the high density disc and leaving the envelope under sampled. In addition to this method of selection, 5\% of the points are linearly distributed in x, y and z with no bias with regards to density or abundance. This provides a minimum level of sampling for the large low density regions in the outer parts of the simulated volume. See fig. \ref{points} for an example of the points distribution in r, z. The function comparing the candidate point to the reference point and the candidate point distribution were selected empirically to sample all scales while ensuring that the majority of points are located in the inner disc where the density is higher.  \smallskip

@@

We focus on the frequency range available with ALMA, with particular attention to band 7, which offers the best trade off between resolution and sensitivity. The results presented in this section are limited to those lines with detectable emission/absorption which can be used to trace either spiral structure or rotation. Thus, we consider C$^{17}$O(3$\rightarrow$2), HCO$^+$(3$\rightarrow$2), OCS (26$\rightarrow$25) and H$_2$CO(4$_{04}\rightarrow$3$_{03}$).  The frequencies and upper level energies of these transitions are given in table \ref{sigmas}. \smallskip

To simulate observations, the model was placed at roughly the distance of nearby low-mass star forming regions (100$\,$pc). The inclination angle was varied from 15$^\circ$ to 75$^\circ$ relative to the edge on case. From these simulated observations, integrated intensity maps, intensity weighted velocity maps and position velocity diagrams were created. The integrated intensity and  intensity weighted velocity maps (figs. \ref{sim_all}, and \ref{mom0_maps}) are integrated from -12.5 to -0.5$\,$km$\,$s$^{-1}$ and +0.5 to +12.5$\,$km$\,$s$^{-1}$ to avoid domination by the contribution from the envelope, which can be seen in some PV diagrams as the strong absorption feature at all positions around zero velocity (see section \ref{sec:alma_predictions}). Intensity weighted velocity maps are shown with a cut-off of 3$\sigma$ as described in section \ref{sec:alma_predictions}.\smallskip

@@

OCS 26$\rightarrow$25 traces only the innermost 20$\,$au of the disc and can be used to examine the central regions (fig. \ref{sim_all}). As OCS is not seen in outflows (e. g. \citealt{Stanke2007}, \citealt{VDTak2003}), it can be used to measure the rotation of the central part of the disc without contamination from shocked material along the outflow. The OCS lines are unique amongst the lines simulated in that they trace radii $<$16$\,$au, even smaller than the extent of the OCS column density distribution shown in fig. \ref{Chemistry}. This is because the OCS line, which has the highest upper energy level among the selected transitions, traces only the hottest and densest regions of the disc, where the high energy states are populated. As a result, the innermost spiral structure is best seen in OCS (26$\rightarrow$25), as shown in fig. \ref{sim_all}. The OCS line is mostly seen in absorption against the bright continuum emission from the disc mid-plane. The exception to this is towards the centre of the disc, where the hole in the hot, dense mid-plane produces little continuum to be absorbed. However, this may be an artefact of our treatment of the inner hole, given that we neglect the presence of the protostar and other hot material within the central 2$\,$au. Thus, we expect absorption also to be observed toward the central region of the disc.\smallskip

The abundance of the H$_2$CO molecule traces the spiral structure of the inner $\sim$40$\,$au of the disc, which can be seen in the integrated intensity map, and its rotation is detected out to larger radii compared to the OCS line (fig. \ref{sim_all}). As with the OCS line, the H$_2$CO 4$_{04}$$\rightarrow$3$_{03}$ line is mostly seen in absorption against the disc mid-plane continuum, with the central region seen in emission. However, unlike the OCS, the H$_2$CO extends out to large enough radii to trace the voids between the outer spiral arms. In these regions, H$_2$CO line emission can be seen.\smallskip

@@

Simulations at inclinations other than 30$^\circ$ were also performed with C$^{17}$O. Fig. \ref{other_inc} shows the C$^{17}$O 3$\rightarrow$2 line in the same physical model but at 15$^\circ$, 30$^\circ$, 45$^\circ$ and 75$^\circ$ to edge on. At 15$^\circ$ to edge on, some structure in the integrated intensity map is still observable and the spiral pattern is clearly seen in the position velocity diagram. At larger inclinations it gets more difficult to obtain any kinematic information but the physical structure of the disc becomes clearer.

@@

\begin{figure*}
 \includegraphics[width=168mm]{Figures/sim/fig11.eps} 
 \caption{H$_2$CO 4$_{04}\rightarrow$3$_{03}$ channel maps as simulated with CASA. The contours start at 0.35 mJy$\,$beam$^{-1}\,$km$\,$s$^{-1}$  and are in steps of 0.35 mJy$\,$beam$^{-1}\,$km$\,$s$^{-1}$ (absorption in blue, emission in red). Overlaid on the 1mm continuum emission. The integration range for each map is 2.4 km$\,$s$^{-1}$, for the top left the range is from -10.8 to -8.4 km$\,$s$^{-1}$ and for the bottom right the range is from 8.4 to 10.8 km$\,$s$^{-1}$.}
 \label{h2co_chanmap}
\end{figure*}

Fig. \ref{continuum} shows the 1\,mm continuum emission as observed with the chosen ALMA configuration. The spiral structure of the disc is clearly visible at this inclination and and it becomes clearer at angles closer to face-on. A mass estimate for the disc was calculated using the equation $M=g S_\nu d^2/\kappa_\nu B_\nu(T_d)$ where $g$ is the gas/dust mass ratio, $S_\nu$ is the integrated flux, $d$ is the distance to the source, $\kappa_\nu$ is the dust opacity and $B_\nu(T_d)$ is the black body function for a given dust temperature \citep{Beltran2006}. Using the above equation and integrating the flux from y=0.2 to -0.2 and from x=0.4 to -0.4, we obtain a disc mass of 0.1 and 0.005$\,$M$_\odot$, if the dust temperature is fixed at 40 and 400\,K, respectively. This shows that the disc is optically thick to millimetre wavelengths and millimetre observations underestimate the disc mass by at least a factor of 4 (see also \citealt{Forgan2013}). In young, massive discs such as those we are investigating, longer wavelength continuum observations are necessary for determining the disc mass.\smallskip

Fig. \ref{h2co_chanmap} shows a channel map of the H$_2$CO 4$_{04}\rightarrow$3$_{03}$ line overlaid on the 1$\,$mm continuum emission. From this it can be seen that the absorption features trace the disc spiral structure whilst the line emission emanates from gaps in the outer disc, where there is less continuum to be absorbed. The central panel is dominated by the envelope contribution, causing strong absorption throughout the disc. The other panels show the line following a rotation pattern distorted by the presence of the spiral arms. \smallskip

Fig. \ref{15deg} shows the CASA simulation of the C$^{17}$O 3$\rightarrow$2 integrated intensity and PV diagram for an inclination angle of 15$^\circ$ to edge on.  Spiral structure can be clearly seen in the position velocity diagram even at low inclinations. The ``finger-like'' structures in the PV diagram are similar to those seen in observations of the Milky Way and in simulations of spirals in the Milky Way (e. g. \citealt{Bissantz2003}). Although the detail of the spiral structure (number of arms, pitch angle etc.) is difficult to gauge from such a PV diagram, high spectral and angular resolution observations can definitely unveil their presence and thus be used to find gravitationally unstable discs. \smallskip

@@

The spectra presented in Fig. \ref{spectra} were extracted from the maps in fig. \ref{mom0_maps} for the positions indicated by the corresponding coloured dots along the Y=0 axis. As with all the CASA simulated images, these spectra were obtained with an ALMA window with total bandwidth 938$\,$MHz, giving velocity resolutions between 418 and 527\,m\,s$^{-1}$. Very notably, these spectral features are not symmetric about the disc centre. For example, C$^{17}$O(3--2) has a significantly stronger blueshifted absorption, due to the presence of a spiral arm along the same line of sight. In the OCS lines, the two sides are not moving with equal velocities and the line widths and shapes are markedly different. These can be attributed to the non-axisymmetric nature of the model.

@@

In this paper we have presented radiative transfer simulations of a hybrid model comprising a 0.39$\, M_\odot$ self gravitating disc with a radius of 64$\,$au and spiral density waves, surrounded by an envelope that is a contracting 10$\,M_\odot$ BE-sphere. The main results of these simulations are as follows:\newline
$\bullet$ CASA simulations of this model show that at a distance of 100$\,$pc, extraction of kinematic and structural information from the continuum and molecular lines is possible with ALMA band 7.\\
$\bullet$ Our simulations show that molecular features are predominantly seen in absorption against the hot mid-plane of self-gravitating protoplanetary discs, with emission only being seen where voids in the structure of the disc do not provide a bright continuum source to be absorbed.\\
$\bullet$ The quiescent nature of the envelope around such discs only affects lines within $\pm\,$0.5 km$\,$s$^{-1}$ of the LSR velocity.\\
$\bullet$ The lines studied (OCS 26$\rightarrow$25, H$_2$CO 4$_{04}$$\rightarrow$3$_{03}$, C$^{17}$O 3$\rightarrow$2 and HCO$^+$ 3$\rightarrow$2) taken together allow all regions of the disc to be sampled and the rotation of the disc to be constrained from the inner edge to the outer edge.\\
$\bullet$ Spiral structures in young embedded discs are detectable at a wide range of inclination angles. They can be resolved spatially at angles close to face on and can be inferred from position velocity diagrams at low inclination angles.\\
{\bf $\bullet$ The dust continuum emission at millimetre wavelengths is optically thick and millimetre observations underestimate the disc mass by significant factors ($>$ 4).\smallskip}

@@

These results demonstrate that ALMA will be able to probe the very earliest stages of circumstellar discs which feed the growing protostar and eventually will evolve in a planetary system. Our results show that ALMA observations in band 7 will allow the assessment of whether or not these young discs are gravitationally unstable and test theories of pre-stellar to protostellar core evolution. If such gravitational instabilities exist, they will have significant implications for the evolution of the disc. For example, they will affect fragmentation, planetesimal formation \citep{Boley2009,Johnson2013,Gibbons2012} and the chemical evolution in the disc (I2011).
